\providecommand{\toplevelprefix}{../..}  %
\documentclass[../../book-main.tex]{subfiles}

\begin{document}

\chapter*{Preface}
% \vspace{-2cm}
\begin{center}
    ``{\em All roads lead to Rome}.''

\end{center}
\vspace{5mm}

\paragraph{Objective.}
This book reveals and studies a common and fundamental problem behind
almost all modern practices of machine (artificial) intelligence.
That is, {\em how to effectively and efficiently learn a
    low-dimensional distribution of data in a high-dimensional space and
    then transform the distribution to a compact and structured
representation?} For any intelligent system, natural or man-made,
such a representation can be generally regarded as a {\em memory} or (empirical)
{\em knowledge} learned from data sensed from the external world. In
recent years, people often informally refer to it as a ``world model.''

\paragraph{Intended Audience.}
This textbook aims to provide a systematic introduction to the
mathematical and computational principles for learning (deep)
representations of such data distributions, as a computable form of
memory, for {\em senior
undergraduate students and beginning graduate students}. The main
prerequisites for this book are undergraduate linear algebra,
probability/statistics, and optimization. Some familiarity with basic
concepts from signal processing (sparse representation and compressed
sensing in particular), information theory, and feedback control
would enhance your appreciation.

\paragraph{Motivation.}
The main motivation for writing this book is that there have been
tremendous developments in the past several years, by the authors and
many colleagues, that aim to establish a principled and rigorous
approach to understand deep neural networks and, more generally,
intelligence itself. The deductive methodology advocated by this new
approach is in direct contrast, and highly complementary, to the
dominant methodology behind current practices of artificial
intelligence, which is largely inductive and trial-and-error. The
lack of understanding about such powerful AI models and systems has
led to increasing hype and fears in society. We believe that a
serious attempt to establish a principled approach to understand
intelligence is more needed than ever. An overarching goal of this
book is to provide solid theoretical and experimental evidence
showing that it is now possible to study intelligence as a scientific
and mathematical subject. As we will argue that intelligence is the
fundamental capability to  develop new memory (or knowledge) or
correct existing one. Hence, one may view this book as a first
attempt to develop {\em a Mathematical Theory of Intelligence}, at
the level of learning empirical knowledge as memory, as the subtitle
of the book suggests.

At the technical level, the theoretical framework presented in this
book helps reconcile a long-standing gap between the classical
approach to modeling data structures that are mainly based on
analytical geometric, algebraic, and probabilistic models (e.g.,
subspaces, Gaussians, and equations) and the ``modern'' approach that
is based on empirically designed non-parametric data-driven models (e.g., deep
networks). As it turns out, a unification of the two seemingly
separate methodologies becomes possible and even natural if one
realizes that they all try to learn and represent {\em low-dimensional}
structures in the data distribution of interest. They are merely
different ways to pursue, represent, and exploit the low-dimensional
structures. From this perspective, even many seemingly unrelated
computational techniques, developed independently in separate fields
at different times, can now be better understood under a common theoretical and
computational framework and probably can be studied together from now
on. As we will see in this book, these techniques include but are not
limited to:
\begin{enumerate}
    \item Lossy compressive encoding-decoding developed in classic
        information theory and coding theory versus modern
        representation learning;
    \item Denoising and dimensionality reduction in classical signal
        processing versus diffusion and denoising models for modern
        generative methods;
    \item Bayesian inference via maximum a posteriori or conditioned
        sampling versus  constrained optimization via continuation
        techniques\footnote{such as the augmented Lagrangian method}.
\end{enumerate}

\paragraph{Main Content.}
We believe that the unified conceptual and computational framework
presented in this book will be of great value to readers who truly
want to clarify mysteries and misunderstandings about deep neural
networks and (artificial) intelligence. Furthermore, the framework is
meant to provide readers with guiding principles for developing
significantly better and truly intelligent systems in the future.
More specifically, besides an informal introduction (chapter), the
main technical content of the book will be organized as six closely
related topics (chapters):
\begin{enumerate}
    \item We will start with the classical and most basic models of
        Principal Component Analysis (PCA), Independent Component
        Analysis (ICA), and Dictionary Learning (DL), which assume
        that the low-dimensional distributions of interest have
        linear and independent structures. From these simple
        idealistic models that are well studied and understood in
        signal processing and compressed sensing, we will introduce
        the most basic and important ideas\footnote{such as denoising
        and dimensionality reduction} for how to learn and represent
        low-dimensional distributions.

    \item To generalize these classical models and their solutions to
        low-dimensional distributions that may not be easily
        represented any simple analytic models, we introduce a
        universal computational principle for learning such
        distributions: {\em compression}. As we will see, data
        compression provides a unifying view of all seemingly
        different classic and modern approaches to distribution or
        representation learning, including entropy minimization,
        denoising via score-matching,  lossy compression with rate
        distortion, and discriminative representation learning via maximizing
        information gain.

    \item Within this unifying framework, modern Deep Neural Networks
        (DNNs), such as ResNet, CNN, and Transformer, can all be
        mathematically interpreted as (unrolled) optimization
        algorithms that iteratively achieve better compression and
        better representations by reducing coding length/rate or
        gaining information. Not only does this framework help
        explain empirically designed deep network architectures thus
        far, it also leads to new architecture designs that can be
        significantly simpler and more efficient.

    \item Furthermore, to ensure that the learned representation for
        a data distribution is correct and consistent, the {\em
        auto-encoding} architectures that consist of both encoding
        and decoding become necessary. In order for a learning system
        to be fully automatic and continuous, we will introduce a
        powerful {\em closed-loop transcription framework} that
        enables an auto-encoding system to self-correct and thus
        self-improve via a minimax game between the encoder and decoder.

    \item To connect theory to practice, we will then study how the
        learned data distribution and representation can be utilized
        as a powerful prior to conduct Bayesian inference or
        constrained optimization that manifests as almost all types
        of tasks and settings that are popular in the practice of
        modern artificial intelligence, including conditional
        estimation, completion, and generation of real-world
        high-dimensional data such as images and texts.

    \item Last but not least, we will demonstrate step-by-step how to
        effectively and efficiently learn deep representations of
        low-dimensional data distributions with large-scale
        real-world datasets, including visual data, body
        motions, and text data, and use them in many practical
        applications such as image classification, image completion,
        image segmentation,
        image generation, motion estimation, and similar tasks for
        the text data.
\end{enumerate}

\paragraph{Summary.}
To summarize, the technical content presented in this book
establishes strong conceptual and technical connections between the
classical analytical approach and the modern data-driven approach,
between simple parametric models and deep non-parametric models,
between diverse inductive practices and a unified deductive framework
from first principles. We will reveal that many seemingly unrelated
or even competing approaches, though developed in separate fields
with different terminologies and at different times in history, yet
all strive to achieve a common objective:
\begin{quote}
    \centering{\em pursuing and exploiting intrinsic low-dimensional
    structures of data distributions embedded in high-dimensional spaces.}
\end{quote}
To this end, the book will take us through a complete journey from
theoretical formulation to mathematical deduction, then to
computational implementation and practical applications.

\end{document}
